\begin{abstract}
Small language models are making edge intelligence practical: a compact model can now classify, triage, and interact with local operational data without sending every query to a cloud API. In industrial Internet of Things (IIoT) security, this shift is attractive because gateways observe sensitive and heterogeneous telemetry, but the model must continue improving as attacks and device behaviors evolve. Federated learning offers a path for collaborative evolution, yet raw model updates can still leak information, and client-side differential privacy (DP) can severely degrade adapter utility. We propose \method, an auditable framework for privately federating a Qwen3.5-2B edge security assistant. \method constructs adapter bases only from public backbone spectra, trains tiny layer-wise core matrices, and releases clipped and noised core updates under client-level DP. To reduce utility loss, \method couples public spectrum-driven rank allocation, layer-adaptive clipping and noise allocation, local residual personalization, and privacy-cost-free server post-processing. We evaluate the framework on \dataset using a mainstream 80/10/10 stratified closed-set protocol and simulated federated client partitions on one server. The planned evaluation compares against FedAvg-LoRA-DP/DP-LoRA-style, a Fed-SB-style fixed-core baseline, non-DP central/local adapters, classical tabular IDS models, and detailed ablations under matched privacy, communication, and trainable-parameter budgets. \todo{Replace this planning abstract with result-backed claims after experiments are complete.}
\end{abstract}
