\paragraph{Edge small language models and on-device intelligence.}
Recent edge-AI surveys argue that on-device models are increasingly motivated by real-time processing, resource constraints, privacy, and reduced dependence on cloud infrastructure~\citep{wang2025ondevice}. Small language models extend this trend from classification to more interactive and task-oriented intelligence: they can support local interpretation, operator-facing interaction, and domain-specific task execution. Our work follows this direction but focuses on the missing training-time question: how can an edge small language model continue improving across private sites without centralizing telemetry or relying on API-based learning?

\paragraph{Federated learning and private adaptation.}
Federated averaging enables decentralized training without moving raw data to a central server~\citep{mcmahan2017fedavg}. In security and IIoT applications, this is attractive because traffic logs and device telemetry can be sensitive. However, model updates can still leak information, motivating formal DP mechanisms~\citep{dwork2006dp,abadi2016dpsgd,mironov2017rdp}. This work uses client-level release DP as the primary guarantee: each participating client clips and noises its upload before transmission. This is stricter than relying only on a trusted server or secure aggregation for central DP, but it can impose a larger utility cost.

\paragraph{Parameter-efficient language model fine-tuning.}
LoRA freezes a pretrained model and learns low-rank update factors, reducing trainable parameters and memory relative to full fine-tuning~\citep{hu2021lora}. Several follow-up methods improve initialization, rank allocation, or adapter efficiency; PiSSA, for example, initializes adapters from principal singular components of pretrained weights~\citep{meng2024pissa}. In federated settings, directly averaging low-rank factors can be problematic because different factor pairs may represent similar products. Fixed-basis or square-core approaches reduce this ambiguity by learning a small matrix between fixed factors.

\paragraph{Differentially private federated PEFT.}
DP-LoRA and DP-DyLoRA study differentially private federated low-rank adaptation for large transformer models and show why PEFT is useful when DP noise makes full-model updates impractical~\citep{liu2023dplora,xu2024dpdylora}. We use these works to define a controlled FedAvg-LoRA-DP baseline under the same client-level accountant, communication reporting, and Qwen3.5-2B backbone. \method differs by moving from ordinary LoRA coordinates to public-spectral core coordinates and by allocating rank and noise with public backbone spectra.

\paragraph{Fed-SB and collision boundary.}
Fed-SB is the closest methodological baseline for this paper~\citep{singhal2025fedsb}. It uses LoRA-SB to learn a small square matrix between fixed adapters, making direct averaging effective and communication efficient, and it studies private federated fine-tuning. Therefore, this paper should not claim that training a square core or exact core averaging is novel. \method differs in how the basis and budgets are chosen: bases are constructed from public backbone spectra rather than task-specific private updates, and layer-wise ranks and privacy noise are allocated under explicit edge and privacy budgets.

\paragraph{IIoT intrusion detection datasets and baselines.}
\dataset provides a labeled IoT/IIoT cybersecurity benchmark for centralized and federated learning~\citep{ferrag2022edgeiiotset}. ToN-IoT and WUSTL-IIOT-2021 are also widely used IIoT security benchmarks and are natural appendix candidates for external validation~\citep{alsaedi2020toniot,zolanvari2019iiot,wustl2021iiot}. Classical tabular models such as Random Forest, XGBoost, LightGBM, MLPs, and transformer-style tabular models remain strong baselines for intrusion detection. We include these baselines to make clear that the target is not simply replacing tabular classifiers, but studying a private federated edge assistant trade-off that combines language-model adaptation, communication efficiency, and formal privacy accounting.
