\paragraph{From compression to evolving edge intelligence.}
The larger motivation is not only to make fine-tuning cheaper. As small language models become more capable, edge deployments can host assistants that classify, summarize, and interact with local operational data. The remaining bottleneck is how such assistants improve after deployment without turning every gateway into a source of cloud API telemetry. \method treats private federated adaptation as the learning mechanism for this edge-intelligence setting.

\paragraph{What the method should be credited for.}
The strongest claim is not that a square core is new. The strongest claim is that public backbone spectra can provide a private-data-free adaptation subspace and a public signal for rank and privacy-noise allocation, thereby improving the utility-privacy-communication trade-off under client-side release DP.

\paragraph{Why client-side DP matters.}
Client-side release DP is more conservative than central DP with a trusted aggregator. It is also more aligned with an honest-but-curious server that can inspect individual uploads. The expected trade-off is lower raw utility than non-DP or trusted-aggregator settings, which makes utility recovery a central contribution rather than an implementation detail.

\paragraph{Limitations.}
The current protocol simulates FL clients on one server and uses a mainstream row-level split as the primary \dataset protocol. This is appropriate for comparability but does not fully test deployment heterogeneity, time shift, unseen attacks, or real gateway scheduling. Record-level privacy is not claimed unless local training is changed to DP-SGD. Finally, language-model adapters should be interpreted as an edge security assistant path, not as a claim that LLMs universally dominate specialized tabular IDS models.

\paragraph{Failure modes to report.}
The final paper should explicitly report cases where adaptive allocation hurts, where residual personalization improves local F1 but degrades global test performance, where clipping rates are too high or too low, and where rare attack classes remain difficult under low $\epsilon$.
